Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-09 write-ups/PLDI/paper.tex
Progress.

@@

\ADComment{Address issue of placing constructs manually --
  transactional semantics and barriers.  Argue why it's easy to do this.}

\ADComment{Mention dynamic parallelism ane explain why it's no good.}

@@

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
As another example, work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.  \ADComment{Perhaps
  make the point that inter-workgroup synchronization is possible in
  very rare cases without blocking, e.g. threadfence reduction.}

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  \ADComment{Shorten} Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex and workgroup $B$ is
spin-waiting for the mutex to be released, an unfair scheduler can
forever schedule $B$, so that $A$ never releases the mutex.
Similarly, an unfair scheduler can induce starvation if workgroup $A$
is at an inter-workgroup barrier awaiting the arrival of other
workgroups, by allowing A to spin forever so that the other workgroups
do not progress to the barrier.

\paragraph{Occupancy-bound execution}
\ADComment{Add footnote to acknowledge multiple wgs per cu}
\ADComment{Say that current GPU schedulers are not required to be fair.}
Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
specify that workgroups must be fairly scheduled, and implementations
of these programming models do not provide fair scheduling in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is via an \emph{occupancy-bound} execution model, whereby
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  Experiments suggest that
the occupancy-bound execution model is widely implemented by today's
GPU devices and drivers~\cite{...}, and it is evidently a simple model
for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model, which they exploit by scheduling no more workgroups
than there are available compute units~\cite{...}.  This works
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  \ADComment{Could cut or move to later:} Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering.  If this watchdog is disabled,
execution of a long-running kernel causes the display to freeze for
the duration of kernel execution.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that
\textbf{(1)} GPU vendors will not commit to the occupancy-bound
execution model that they currently implement, due to the need for
multi-tasking and energy throttling, yet \textbf{(2)} GPU vendors will
not guarantee fair scheduling, due to the runtime overhead associated
with supporting full preemption of workgroups~\cite{ISCAPAPER}.  Vendors instead wish to retain the essence of the
occupancy-bound execution model, which is simple to implement
efficiently, and provide preemption for \ADComment{shorten} the special cases associated
with multi-tasking for high-priority jobs (most crucially graphics)
and energy throttling.  In particular, they do not want to commit to
fair scheduling of a workgroup once it has been preempted.

\ADComment{Kepler preemption support?}

\paragraph{Our proposal}
%
To summarise: blocking algorithms, such as most irregular algorithms,
demand a degree of fair scheduling of workgroups, but for good reasons
GPU vendors will not commit to the guarantees of even the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and written using two additional language
primitives, $\offerkill$ and $\offerfork$, which must be placed
carefully by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$ to sacrifice itself to
the scheduler.  This indicates to the scheduler that the workgroup
would ideally continue with its execution, but that if required the
scheduler may destroy the workgroup for good: the cooperative kernel
is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates to the scheduler that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commencing execution at the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel, with
both functional and non-functional requirements.  Functionally, the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled, while the cooperative kernel must be robust to
workgroups leaving and joining the computation in response to $\offerkill$
and $\offerfork$.
Non-functionally, a cooperative kernel must ensure that $\offerkill$
is executed frequently enough that the scheduler can accommodate
soft-real time constraints, e.g.\ allowing a smooth frame-rate for
graphics or allowing compute units to be powered down quickly when
required.  The scheduler should respond to $\offerkill$ and
$\offerfork$ calls in a manner that avoids under-utilisation of
hardware resources by the cooperative kernel, e.g.\ the scheduler
should not kill workgroups unless they are required for another
purpose.

Cooperative kernels have several appealing properties:

@@

\item They
accommodate the needs of the irregular
data-parallel algorithms that have been successfully accelerated using GPUs: such algorithms can be implemented naturally
as cooperative kernels \ADComment{boast that we did a good set}, so
that fair scheduling between workgroups is guaranteed.

\item They
have no impact on the development of regular (non-cooperative
kernels): these can be programmed exactly as they are now.

\item The
programming model is backwards compatible: a non-blocking algorithm
implementation that works on today's GPU platforms can be upgraded to
use our language extensions for cooperative scheduling, and will
behave exactly as it does at present if the new $\offerkill$ and
$\offerfork$ functions are simply ignored.

\item The programming
model is easy to implement on top of the occupancy-bound execution
model that current GPU platforms provide; we present a
proof-of-concept implementation on top of OpenCL 2.0 that requires no
special-purpose hardware or driver support.

\item Our results
suggest that cooperative kernels can provide efficient multi-tasking
of cooperative and non-cooperative tasks: our experiments across a
range of GPUs from two vendors show very good response times for
scheduling short, non-cooperative kernels concurrently with
long-running cooperative kernels.  These performance results are with
respect to our proof-of-concept, portable implementation; a vendor
specific implementation of cooperative kernels could likely yield
better performance still through device-specific driver and compiler
support.

@@

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms  \ADComment{Maybe big up to say that we give semantics to the new operations.}

\item A portable implementation of cooperative kernels on top of the OpenCL 2.0 programming model

\item The porting of \ADComment{TODO} irregular graph algorithms \ADComment{tighten} to the cooperative kernels model

\item \ADComment{The experiments}

\ADComment{Text to maybe work in.}

In traditional co-operative multi-tasking, a task relinquishes control
and regains control in a stateful manner. We propose a lighter weight
model we believe to suitible for the type of blocking algorithms we
are aware of for GPUs.

The scheduler agrees to scheduler the workgroups fairly, but is at
liberty to eliminate workgroups for good when they offer themselves
to be killed. This allows the schduler to repond to changing depands
on the GPU, e.g. to implement energy throttling.

Prototype implementation of the programming model using OpenCL 2.0
advance features to build the scheudler and implement a clang based
compiler to transform input programs using the language extensions to
lower-level programs that interact with the scheduler. Our prototype
simulates multiple independent kernels into a mega-kernel. A vendor
specific implementation could achieve higher efficiency through
modications to the GPU driver that may take advantage of hardware
features. (e.g. mega kernel might not be good to compile). Its
reasonable to think our proof of concept provides a lower bound on how
well we can do.

We ported a large set of blocking algorithms to our programming
model. We evaluate co-operative scheduling with a cooperative kernel
and non-coperative kernel. GPUs from multiple vendors and access
various things.

Our results are promising.

In the evaluation, make a note that the applications we worked on
Namely if they use only a barrier or work-stealing idiom, are
straightforward to adopt to our programming model.

\section{Motivating example: work stealing}

@@

\section{The Cooperative Kernels Programming Model}

We describe our cooperative kernels programming model as an extension
to the industry-standard OpenCL programming model for heterogenous
many-core architectures.  However, the cooperative kernels concept is
more general, and could be applied to extend other GPU programming
models, such as CUDA and HSA.  \ADComment{However, unclear that we could implement it on top of these, but that's not really the point.}

\subsection{Semantics of Cooperative Kernels}

@@

\item for each variable $v$ annoated with \restore, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{ }, an inter-workgroup barrier
requires fair scheduling of workgroups.

@@

The $\globalbarrier()$ primitive can be implemented directly using the
atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
but the implementation is involved, hence why we include this function
as a primitive.  As noted above, the $\resizingglobalbarrier()$
primitive is, semantically, synactic sugar for a sequence of other
primitive calls.  We include it as a primitive because (a) we have
found that global barriers often provide a natural point for
cooperation with the scheduler, and (b) a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described
above can be made significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in Section~\ref{...}.

\subsection{Programming guidelines}

@@

The programer is responsible for keeping at least one workgroup alive.  In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.  This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:

@@

If a cooperative kernel does not directly issue $\offerkill$ and
$\offerfork$ calls then the only possibilty for the number of
workgroups to change is due to $\resizingglobalbarrier$ being invoked.

\ADComment{Shorten sentences.}
In this case, because the resizing barrier is a kernel-level function,
the statements executed by the cooperative kernel can be divided into
\emph{resizing barrier intervals}: a resizing barrier interval starts
at a resizing barrier or the kernel entry point, and ends at a
resizing barrier or a kernel exit point.  At any moment of execution,
all threads are guaranteed to be executing within the same resizing
barrier interval.  As a result, the number of workgroups executing the
kernel is guaratneed to remain unchanged within each resizing barrier
interval, so the result returned by $\getnumgroups{0}$ is guaratneed
to be identical on repeated calls during such an interval.

This guarantee may be exploited by algorithms that peform strided data
processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?}

\subsection{Non-functional Requirements of Cooperative Kernels}

The semantics presented in Section~\ref{...} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.

@@

To achieve this balance, both an implementation of the ooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet a number of non-functional requirements, which we outline
here.

@@

\subsection{Alternative Semantic Choices}

@@

\ADComment{What are difficultiese with other alternatives and why we didn't use them.}

@@

\subsection{Backwards Compatibility}

@@

\subsection{Preemption-aware work stealing}

\ADComment{Remind the reader that these are inspired by work stealing
  and barrier synchonization.  In \mysec\ref{ } we describe the
  semantics of kernels that use these primitives.  In \mysec\ref{ } we
  show how work stealing can be implemented.  In \mysec\ref{ } we show
  how a barrier can be implemented.}

\ADComment{Show how work stealing can now be made safe.}

\subsection{A preemption-aware global barrier}

\ADComment{Show how the global barrier can be written in a manner that
  maintains contiguous ids using the refined semantics.}

\section{Application-specific optimisations}

\ADComment{Explain that for global barriers, we explain how some new
  primitives can lead to a better implementation.  Explain that for
  work stealing there are various ways the existing ckill and cfork
  primitives can be used.}

\subsection{An optimised barrier implementation}

\ADComment{Remark that the barrier of \mysec\ref{sec:naivebarrier} may
  be inefficient, discuss the use of query primitives to make this
  more efficient.}

\subsection{Variations on work stealing}

\ADComment{Three cases here: always ckill/cfork, ckill/cfork only when
  queue is empty, and have a ``mediator'' workgroup to manage things.}

\section{A mega kernel-based prototype}

@@

\section{Experimental evaluation}

\ADComment{TODO: write up notes from whiteboard chat.}

\section{Related work}

\section{Conclusions and future work}

@@

\section{Old intro}

\ADComment{Points to make: It is important to be able to execute
  long-running, cooperative applications on GPUs (give examples).
  However, resource sharing and resource flexibility is also important
  (give examples, including sharing resources with graphics, sharing
  resources between compute taks, and being flexible about resources
  under energy constraints), and these are not supported by today's
  GPU platforms: long-running applications cause display freezes, etc.
  Resource-based preemption, whereby workgroup execution is postponed
  to free up resources for other purposes (or to reduce resources to
  save energy) can potentially enable resource sharing.  There is
  evidence in the research literature that efficient resource sharing
  for GPUs is possible (cite academic works), and signs from industry
  that this is coming in future architectures.  [Explain more clearly
    what we mean by resource-based preemption.]  Unfortunately,
  arbitrary resource-based preemption is fundamentally incompatible
  with cooperative applications---only trivial examples of cooperation
  are safe [reduction, load balancing via atomic increment]; the HSA
  model of guaranteeing preemption in reverse order of IDs allows some
  cooperative applications to be implemented.  Still, in the general
  case even a mutex cannot be used safely.}

\ADComment{Say that we believe the problem can only be solved through
  programming language innovations with appropriate runtime support.
  We have observed inter-workgroup barriers and work stealing to be
  the main use cases for cooperation, and so in this paper propose two
  new language primitives, ckill and cfork, designed to meet the needs
  of these primitives.  [Describe them a bit.]}

\ADComment{Move on to our main contributions}

\ADComment{This is what we had before.}

Graphics processing units (GPUs) are highly parallel co-processors that are now incorporated into many
devices, ranging from the world's most powerful supercomputers to battery-powered devices such as tablets and mobile phones. While
originally designed to accelerate graphics applications, GPUs are now
used in a wide range of applications, including molecular simulations
and weather forecasting. \ADComment{Expand the list of application areas a bit more; mention their attractiveness due to energy-efficiency.}

New application domains for GPUs are enabled by general purpose
programming languages with associate compilation tools that target GPU architectures.
The current most popular GPU programming languages are CUDA and OpenCL. These languages allow developers to
write C-like programs that can be dispatched and executed on a GPU
device.

GPUs are massively parallel, employing many threads to execute a
program.  These threads are organised hierarchically: groups of threads
are called \emph{workgroups}, and threads interact differently depending on whether they reside in the same workgroup or not.  Traditionally, applications accelerated by GPUs are
data-parallel, requiring little interaction between threads. A small
set of bulk synchronisation primitives are provided and any
inter-thread communication must be performed around these primitives.  \ADComment{Do we want to make the point that this is traditionally limited to the intra-workgroup case?}

While this traditional way to program GPUs has proved useful, it also
has limitations for some applications, in particular applications that
expose a dynamic amount of parallelism throughout execution. Such
applications are said to have irregular parallelism (we call such
applications \emph{IP applications}). Efficient parallel computation
of IP applications requires finer-grained synchronisation than that
which is provided by GPU language primitives.

>>>>

\newcommand{\NumAlgorithms}{8}

@@

\ADComment{Open to attack: why not just buy another GPU?}

@@

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
As another example, work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
another workgroup $B$ to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can a workgroup to spin-wait
at a global barrier so that other workgroups do not make progress
towards the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{...}, CUDA~\cite{...}  and
HSA~\cite{...}---specify almost no guarantees regarding scheduling of
workgroups, and implementations of these programming models do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{...}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{...}.  It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model.  They exploit the model by scheduling no more workgroups
than there are available compute units~\cite{...}.  This works
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering~\cite{...}.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{...}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that (1)
GPU vendors will not commit to the occupancy-bound execution model
that they currently implement, due to the need for multi-tasking and
energy throttling, yet (2) GPU vendors will not guarantee fair
scheduling, due to the runtime overhead associated with supporting
full preemption of workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors
instead wish to retain the essence of the occupancy-bound execution
model, which is simple to implement efficiently, and provide
support for preemption only in key special cases.

\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms, such as most irregular algorithms,
demand a degree of fair scheduling of workgroups, but for good reasons
GPU vendors will not commit to the guarantees of even the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and is written using two additional
language primitives, $\offerkill$ and $\offerfork$, which must be
placed carefully by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates to the scheduler
that the workgroup would ideally continue with its execution, but that
if required the scheduler may destroy the workgroup for good: the
cooperative kernel is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates to the scheduler that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel, with
both functional and non-functional requirements.  Functionally, the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled, while the cooperative kernel must be
robust to workgroups leaving and joining the computation in response
to $\offerkill$ and $\offerfork$.  Non-functionally, a cooperative
kernel must ensure that $\offerkill$ is executed frequently enough
that the scheduler can accommodate soft-real time constraints,
e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
units to be powered down quickly when required.  The scheduler should
respond to $\offerkill$ and $\offerfork$ calls in a manner that avoids
under-utilisation of hardware resources by the cooperative kernel,
e.g.\ the scheduler should not kill workgroups unless they are
required for another purpose, and should facilitate forking of
additional workgroups when possible.

The cooperative kernels programming model has several appealing
properties:

@@

\item By providing fair scheduling between workgroups, cooperative
  kernels meet the needs of blocking algorithms, including irregular
  data-parallel algorithms.

\item The model has no impact on the development of regular
  (non-cooperative) kernels: these can be programmed exactly as they
  are now, as can graphics shading routines written using APIs such as
  OpenGL, Vulkan and DirectX.

\item The model is backwards-compatible: a non-blocking algorithm
  implementation that works on today's GPU platforms can be upgraded
  to use our language extensions for cooperative scheduling, and will
  behave exactly as it does at present if the new $\offerkill$ and
  $\offerfork$ functions are simply ignored.

\item The model can be implemented on top of the occupancy-bound
  execution model that current GPU platforms provide---we present a
  proof-of-concept implementation on top of OpenCL 2.0 that requires
  no special-purpose hardware or driver support.

\item Our experiments across four GPUs from two vendors show that our
  portable, proof-of-concept implementation of cooperative kernels can
  provide efficient multi-tasking of cooperative and non-cooperative
  tasks.  A GPU vendor implementation could likely provide even better
  efficiency, by leveraging platform-specific hardware, driver and
  compiler features.

@@

Placing the $\offerkill$ and $\offerfork$ primitives requires manual
effort and care, but our experience porting a representative set of
GPU-accelerated irregular algorithms to use cooperative kernels
suggests that this is straightforward in practice.  We discuss this in
detail in Section~\ref{sec:portingalgorithms}.

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.  We describe the semantics of the programming model in detail (Section~\ref{sec:cooperativekernels}).

\item A proof-of-concept portable implementation of cooperative
  kernels on top of the OpenCL 2.0 programming model
  (Section~\ref{sec:implementation}).

\item An experience report porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  assessing the performance overhead associated with cooperation and
  the responsiveness to high priority tasks that cooperation enables,
  across four GPUs from two vendors (Section~\ref{sec:experiments}).

We begin by providing an overview of the OpenCL programming model and
two motivating examples of irregular algorithms (Section~\ref{sec:background}).  At the end of the paper we discuss related work (Section~\ref{sec:relatedwork}) and avenues for future work (Section~\ref{sec:conclusion}).

\section{Background and Examples}\label{sec:background}

@@

\section{The Cooperative Kernels Programming Model}\label{sec:cooperativekernels}

We describe our cooperative kernels programming model as an extension
to the industry-standard OpenCL programming model for heterogeneous
many-core architectures.  However, the cooperative kernels concept is
more general, and could be applied to extend other GPU programming
models, such as CUDA and HSA.

We describe the semantics of cooperative kernels
(Section~\ref{sec:semantics}), then summarise the important
differences between programming with regular vs.\ cooperative kernels
(Section~\ref{sec:programmingguidelines}) and the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the applications that we have studied
(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
have taken particular semantic decisions where other reasonable
alternatives were possible (Section~\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(Section~\ref{sec:backwardscompatibility}).

\subsection{Semantics of Cooperative Kernels}\label{sec:semantics}

@@

\item for each variable $v$ annotated with \restore, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{sec:intro}, an inter-workgroup barrier
requires fair scheduling of workgroups.

@@

The $\globalbarrier()$ primitive can be implemented directly using the
atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
but the implementation is involved, hence why we include this function
as a primitive.  As noted above, the $\resizingglobalbarrier()$
primitive is, semantically, syntactic sugar for a sequence of other
primitive calls.  We include it as a primitive because (a) we have
found that global barriers often provide a natural point for
cooperation with the scheduler, and (b) a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described
above can be made significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in Section~\ref{sec:resizingbarrier}.

\subsection{Programming Guidelines}\label{sec:programmingguidelines}

@@

The programmer is responsible for keeping at least one workgroup alive.  In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.  This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:

@@

If a cooperative kernel does not directly issue $\offerkill$ and
$\offerfork$ calls then the only possibility for the number of
workgroups to change is due to $\resizingglobalbarrier$ being invoked.

\ADComment{Shorten sentences.}
In this case, because the resizing barrier is a kernel-level function,
the statements executed by the cooperative kernel can be divided into
\emph{resizing barrier intervals}: a resizing barrier interval starts
at a resizing barrier or the kernel entry point, and ends at a
resizing barrier or a kernel exit point.  At any moment of execution,
all threads are guaranteed to be executing within the same resizing
barrier interval.  As a result, the number of workgroups executing the
kernel is guaranteed to remain unchanged within each resizing barrier
interval, so the result returned by $\getnumgroups{0}$ is guaranteed
to be identical on repeated calls during such an interval.

This guarantee may be exploited by algorithms that perform strided data
processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?}

\paragraph{Placing the new constructs}

\ADComment{TODO: needs work.}
.  The irregular
algorithms we encountered either employ work stealing or operate on
graph data structures.  Work stealing algorithms have a transactional
flavour, whereby tasks are repeatedly pulled from a queue and
processed by a workgroup.  The point of task completion---the end of a transaction---is a natural place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to indicate that

\subsection{Non-Functional Requirements of Cooperative Kernels}\label{sec:nonfunctional}

The semantics presented in Section~\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.

@@

To achieve this balance, both an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet a number of non-functional requirements, which we outline
here.

@@

\subsection{Alternative Semantic Choices}\label{sec:semanticalternatives}

@@

\ADComment{What are difficulties with other alternatives and why we didn't use them.}

@@

\subsection{Backwards Compatibility}\label{sec:backwardscompatibility}

@@

\section{Prototype Implementation}\label{sec:implementation}

@@

\ADComment{Give details of how host and device interact.}

\ADComment{Explain kernel merge tool.}

\subsection{An Efficient Resizing Barrier}\label{sec:resizingbarrier}

\ADComment{Optimised barrier implementation.}

\section{Applications and Experiments}\label{sec:experiments}

\subsection{Porting Irregular GPU Algorithms to Cooperative Kernels}\label{sec:portingalgorithms}

\ADComment{Among other things, make sure we address the point that placing the primitives was pretty easy.}

\section{Related Work}\label{sec:relatedwork}

\section{Conclusions and Future Work}\label{sec:conclusion}

@@

\section{Notes}

  \ADComment{Perhaps
  make the point that inter-workgroup synchronization is possible in
  very rare cases without blocking, e.g. threadfence reduction.}

\ADComment{If this watchdog is disabled,
execution of a long-running kernel causes the display to freeze for
the duration of kernel execution.}

\ADComment{Kepler preemption support?  Mention in related work?}

\ADComment{Mention dynamic parallelism and explain why it's no good.}

\section{Remainder of old intro}
